---
title: "Data Science and Society using R: Homework Template"
author: "Jun Yan"
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    toc-depth: 2
    embed-resources: true
    code-fold: true
    theme: cosmo
---

## Instructions

This is a template for your homework assignments in the *Data Science
and Society using R* course. Please follow the structure below to
complete your assignment. Replace placeholder text (e.g., "[Your
answer here]") with your responses. Ensure all R code is executable
and produces the expected output. Submit your work as a single `.html`
file rendered from this `.qmd` file.


### Submission Guidelines
1. **Complete the Assignment**: Fill in all sections with your
   answers, R code, and explanations.
2. **Render to HTML**: Use Quarto to render this `.qmd` file to
   HTML. In RStudio, click the "Render" button or run `quarto render
   <filename>.qmd` in the terminal.
3. **Check Output**: Ensure the rendered HTML file includes all text,
   code outputs, and visualizations. If there are things to fix, fix
   them in the source and go to Step 2.   
4. **Submit**: Print the HTML output to a PDF file and put it into a
release.

If you encounter issues, consult the
[Quarto documentation](https://quarto.org) or ask around for help.


## Problem 1: Data Exploration

### Question
Load a dataset and provide a summary of its structure and key statistics.

### Example
Below is an example using the built-in `mtcars` dataset in R.

```{r}
# Load dataset
data(mtcars)

# Display structure
str(mtcars)

# Summary statistics
summary(mtcars)
```

### Your Answer
[Load your assigned dataset and provide its structure and summary
statistics here. Include a brief narrative explaining what the dataset
contains and highlight one interesting finding from the summary.]


## Problem 2: Data Visualization

### Question
Create a visualization to explore a relationship between two variables in the dataset.

### Example
Here’s an example of a scatterplot using `ggplot2` to explore the
relationship between `mpg` (miles per gallon) and `hp` (horsepower) in
the `mtcars` dataset.


```{r}
# Load ggplot2
library(ggplot2)

# Create scatterplot
ggplot(data = mtcars, aes(x = hp, y = mpg)) +
  geom_point() +
  labs(title = "Scatterplot of MPG vs Horsepower",
       x = "Horsepower (hp)",
       y = "Miles per Gallon (mpg)") +
  theme_minimal()
```

### Your Answer
[Create a visualization for your dataset. Include R code and a brief
explanation of what the plot shows and why you chose this
visualization.]


## Problem 3: Analysis and Interpretation

### Question
Perform a simple analysis (e.g., correlation, regression, or summary
statistics) and interpret the results in the context of data science
and society.


### Example
Here’s an example of calculating the correlation between `mpg` and
`hp` and interpreting it.

```{r}
# Calculate correlation
cor.test(mtcars$mpg, mtcars$hp)
```

**Interpretation**: The correlation coefficient indicates a negative
relationship between horsepower and miles per gallon, suggesting that
cars with higher horsepower tend to have lower fuel efficiency. This
has societal implications, as fuel efficiency impacts environmental
sustainability and consumer costs.


### Your Answer
[Perform your analysis here. Include R code and a narrative
interpreting the results, connecting them to societal or ethical
considerations relevant to the course.]


## Additional Notes
- Ensure all R packages used are installed (e.g.,
  `install.packages("ggplot2")` if needed).
- Comment your R code to explain your steps.
- Use clear, concise writing for narrative sections.
- If you include external data, provide a brief description of its
  source and ensure it’s accessible (e.g., built-in datasets or data
  provided by the instructor).

