# Exploratory Data Analysis (EDA)

Exploratory Data Analysis (EDA) is the practice of using graphs and
numerical summaries to become familiar with a dataset before formal
modeling. The term was popularized by John Tukey in his influential
book *Exploratory Data Analysis* [@tukey1977exploratory],
and it remains a cornerstone
of modern data science. The purpose of EDA is to reveal the main
characteristics of the data, detect unusual cases, check the quality
of measurements, and generate hypotheses that can guide later
modeling. Unlike confirmatory analysis, which tests specific
hypotheses with formal statistical procedures, EDA is open ended and
iterative. Analysts move back and forth between plotting,
summarizing, and transforming data, refining their understanding at
each step. 


@fig-eda-flow shows the data science workflow in a compact
layout, generated directly in R. Compared with embedding static images,
this approach keeps your notes reproducible and editable. The flowchart
emphasizes that exploratory data analysis sits between data
transformation and modeling. It also highlights feedback loops:
visualization can suggest new transformations, and communication can
lead back to data collection.

```{r}
#| label: fig-eda-flow
#| fig-cap: "Data science workflow from design and collection 
#|    to communication, with feedback loops."
#| echo: false
library(DiagrammeR)

grViz("
digraph {
  graph [layout = dot, rankdir = TB]

  node [shape = box, style = filled, fontname = Helvetica]

  A [label = 'Design & collect', fillcolor = '#FFF3E0', color = '#FF9800', fontcolor = '#E65100']
  B [label = 'Import',           fillcolor = '#FFF3E0', color = '#FF9800', fontcolor = '#E65100']
  C [label = 'Tidy',             fillcolor = '#FFF3E0', color = '#FF9800', fontcolor = '#E65100']
  D [label = 'Transform']
  E [label = 'Visualize',        fillcolor = '#FFF3E0', color = '#FF9800', fontcolor = '#E65100']
  F [label = 'Model',            fillcolor = '#E3F2FD', color = '#2196F3', fontcolor = '#0D47A1']
  G [label = 'Publish']
  H [label = 'Communicate',      fillcolor = '#FFF3E0', color = '#FF9800', fontcolor = '#E65100']

  # main downward flow
  A -> B -> C -> D -> E -> F -> G -> H

  # feedback arrows
  H -> A [style=dashed, color=gray50]
  D -> F [style=dashed, color=gray50]
}
")
```

We will use `ggplot2::diamonds` as a running example. It contains about
fifty four thousand round‑cut diamonds with carat, cut, color, clarity,
depth, table, price, and x/y/z dimensions. The dataset is big enough to
be realistic while still rendering quickly in class.

::: callout-tip
**Tip.** Always write your EDA as runnable code. Reproducible EDA makes
your figures defensible and your insights repeatable.
:::

## Questions to consider

When beginning an EDA it is useful to ask two broad questions that
frame all of the work. These questions remind us that analysis is
about patterns, not isolated numbers.

1. **What type of variation occurs within variables?**  
   Variation refers to the tendency of a variable to take different
   values from one observation to another. Even for simple variables
   like eye color or carat weight, no two individuals are exactly the
   same. Understanding the spread and shape of this variation is the
   first task of EDA.

2. **What type of covariation occurs between variables?**  
   Covariation describes the way two or more variables move together
   in a related fashion. For instance, diamond carat and price tend to
   increase together. Examining such relationships gives insight into
   possible explanations and later modeling choices.

### Useful terms

Clear vocabulary helps organize thinking during EDA.

- **Variable.** A measurable characteristic such as carat, cut, or
  price. Variables may be quantitative or qualitative.

- **Value.** The state of a variable when it is measured. Values can
  differ across observations; for example, two diamonds can have
different carat weights.

- **Observation.** A set of measurements recorded on the same unit at
the same time. In `diamonds`, each row is one diamond with values
for all variables.

- **Tabular data.** A collection of values arranged with variables in
columns and observations in rows. Data is called *tidy* when each
value occupies its own cell, making manipulation and visualization
straightforward.

```{r}
library(tidyverse)
glimpse(diamonds)
```

### Variation

Every variable has its own pattern of variation, and these patterns
often reveal important information. The best way to learn about the
variation of a variable is to visualize its distribution.

- For **categorical variables**, variation shows up in the relative
  frequencies of categories.
- For **quantitative variables**, variation appears in the spread and
  shape of the distribution.

```{r}
# variation in cut (categorical)
diamonds |>
  count(cut) |>
  mutate(prop = n / sum(n))
```

```{r}
# variation in carat (numeric)
library(ggplot2)
ggplot(diamonds, aes(carat)) +
  geom_histogram(binwidth = 0.1, boundary = 0, closed = "left") +
  labs(title = "Variation in carat", x = "Carat", y = "Count")
```

### Visualizing distributions

How to visualize a distribution depends on the type of variable.

- **Categorical variables** take one of a small set of levels and are
  usually stored as factors or character vectors in R. Bar charts are
a natural choice for displaying their distributions.

```{r}
ggplot(diamonds, aes(cut)) +
  geom_bar() +
  labs(title = "Distribution of cut", x = NULL, y = "Count")
```

- **Quantitative variables** take on a wide range of values. Their
distributions are best displayed with histograms, density curves, or
boxplots.

```{r}
ggplot(diamonds, aes(price)) +
  geom_histogram(binwidth = 500, boundary = 0, closed = "left") +
  labs(title = "Distribution of price", x = "Price (USD)", y = "Count")
```

### Outliers

Outliers are values that stand apart from the general pattern of the
rest of the data. They may be errors, unusual but valid cases, or the
most interesting part of the dataset. Detecting outliers is an essential
part of EDA, because they can distort summaries and influence models.

- Outliers in **quantitative variables** are extreme values far from the
  bulk of the distribution.
- Outliers in **categorical variables** are rare levels that occur only a
  handful of times.

```{r}
# look at top-priced diamonds
library(dplyr)
diamonds |>
  arrange(desc(price)) |>
  select(carat, cut, color, clarity, price) |>
  head(10)

# scatterplot to reveal unusual points
library(ggplot2)
ggplot(diamonds, aes(carat, price)) +
  geom_point(alpha = 0.3) +
  labs(title = "Carat vs. price with potential outliers",
       x = "Carat", y = "Price (USD)")
```

::: callout-note
**Note.** Not every outlier should be removed. Some may hold the key to
a new insight, while others may be errors worth correcting. Document any
decisions you make.
:::


## Conditional statements

Many insights arise from studying how a variable behaves given the level
of another variable. These conditional relationships are the basis for
understanding covariation.

- **Numeric given categorical.** Compare distributions of a numeric
  variable across categories.
- **Numeric given numeric.** Study scatterplots, possibly with facets,
  to see how the relationship changes across groups.

```{r}
# price conditional on cut
ggplot(diamonds, aes(cut, price)) +
  geom_boxplot(outlier.alpha = 0.2) +
  labs(title = "Price distribution by cut", x = NULL, y = "Price (USD)")

# carat vs. price conditional on clarity
set.seed(1)
d_small <- diamonds[sample(nrow(diamonds), 8000), ]
ggplot(d_small, aes(carat, price)) +
  geom_point(alpha = 0.35) +
  facet_wrap(~ clarity) +
  labs(title = "Price vs. carat by clarity",
       x = "Carat", y = "Price (USD)")
```

::: callout-tip
**Tip.** Faceting is usually clearer than mapping too many aesthetics at
once. Use small multiples to keep comparisons interpretable.
:::

## Good Practices

- Prefer simple, readable graphics over clever ones.
- Keep code blocks short and name intermediate objects clearly.
- Mix visual and tabular summaries; each catches different issues.
- Record every data change. EDA often uncovers fixes that matter later.
- Revisit EDA after modeling; residuals are just another dataset.

