# Tidy Data {# ch-tidy}


```{r}
#| include: FALSE
library(tidyverse)
knitr::opts_chunk$set(eval = TRUE, echo = TRUE)
```


## Introduction

Working effectively with data in R is greatly simplified by the
`tidyverse`, a collection of packages designed for data science. The
`tidyverse` provides a consistent framework for data manipulation,
visualization, and modeling, which helps students learn generalizable
skills rather than package-specific tricks.


A central concept in the `tidyverse` is the` tibble`, a modern
re-imagining of the traditional R data frame. Tibbles keep the familiar
two-dimensional table structure but introduce improvements such as
preserving variable types, supporting list columns, and displaying data
more cleanly in the console. These features make them easier to use in
practice, especially with large datasets.


Finally, the idea of tidy data lies at the heart of the `tidyverse`.
According to Hadley Wickham’s definition, tidy data means each variable
forms a column, each observation forms a row, and each type of
observational unit forms a table. Tidy data creates a standardized
structure that enables smooth use of functions across the `tidyverse`,
reducing the need for ad hoc data reshaping and making analyses more
transparent and reproducible.

## Tibbles

- What is a tibble?
    + Tibbles are data frames
    + But they try and enhance the regular "old" data frame from base R
- To learn more
    + `vignette("tibble")`
  

Tibbles are data frames that enhance the regular "old" data frame from
base R. They keep the same two-dimensional tabular structure but are
designed to be more consistent, predictable, and user-friendly.


### Creating tibbles

There are several ways to create tibbles depending on the source of the
data.

+ From individual vectors

The simplest way is to build a tibble directly from vectors using
`tibble()`. Inputs of length 1 are automatically recycled, and you can
refer to variables you just created:

```{r}
tibble(
  x = 1:5,
  y = 1,
  z = x ^ 2 + y
)
```

+ Converting existing objects

You can convert existing data structures into tibbles with `as_tibble()`:

```{r}
# From a data frame
head(as_tibble(iris))

# From a list
as_tibble(list(x = 1:3, y = letters[1:3]))
```

+ Reading from external files

Packages in the tidyverse ecosystem return tibbles when reading data
from files:

```{r}
# From CSV, TSV, or delimited text file
head(readr::read_csv("data/Chetty_2014.csv"))

# From Excel files
# readxl::read_excel("data.xlsx")
```

+ Reading from databases

You can also obtain tibbles when working with databases using packages
such as `DBI` and `dbplyr`:

```r
library(DBI)
con <- dbConnect(RSQLite::SQLite(), "mydb.sqlite")
tbl(con, "tablename")   # returns a tibble-like object
```


A tribble is a transposed tibble, designed for small data entry in code.
Column headings are defined by formulas that start with `~`:

```{r}
tribble(
  ~x, ~y, ~z,
  #--|--|----
  "a", 2, 3.6,
  "b", 1, 8.5
)
```

The `tibble` and `readr` packages are part of the core tidyverse, so
they load automatically with `library(tidyverse)`. Other packages such
as `readxl` and `dbplyr` belong to the tidyverse ecosystem. They follow
the same principles and return tibbles, but you need to load them
explicitly.


### Features of tibbles

Tibbles make fewer automatic changes than base R data frames:

- They never change the type of inputs (strings are not converted to
  factors).
- They never change variable names.
- They never create row names.

For example:

```{r}
tb <- tibble(
  `:)` = "smile",
  ` ` = "space",
  `2000` = "number"
)
```

These column names would not be valid in base R, but are allowed in a
tibble.



There are two main differences between tibbles and base R data frames:

**Printing**
Tibbles have a refined print method that shows only the
first 10 rows and only as many columns as fit on the screen:

```{r}
tibble(
  a = lubridate::now() + runif(1e3) * 86400,
  b = lubridate::today() + runif(1e3) * 30,
  c = 1:1e3,
  d = runif(1e3),
  e = sample(letters, 1e3, replace = TRUE)
) |> head()
```

This design avoids overwhelming the console when printing large data
frames.

If you need more output, you can adjust options:

- `print(n = , width = )` controls number of rows and columns.
- Global options can be set with:

```r
options(tibble.print_max = n, tibble.print_min = m)
options(tibble.print_min = Inf)     # always show all rows
options(tibble.width = Inf)         # always print all columns
```

**Subsetting**
Most of the subsetting tools we have used so far
generally subset the entire data frame. To pull out just a single
variable or value, we can use `$` and `[[`:
    - `[[` extracts by name or position
    - `$` extracts by name with less typing

```{r}
df <- tibble(
  x = runif(5),
  y = rnorm(5)
)

# Extract by name
df$x
df[["x"]]

# Extract by position
df[[1]]
df[[1,1]]
```


## Tidy up data

Structuring datasets to facilitate analysis is at the core of the
principles of tidy data, as described by Hadley Wickham.

### Tydy data

Tidy data follows three basic rules:

- Each variable forms a column
- Each observation forms a row
- Each type of observational unit forms a table

When these rules are not followed, the dataset is considered untidy.
Common signs of untidy data include:

- Column headers are values instead of variable names
- Multiple variables are stored in one column (for example, City_State)
- Variables are stored in both rows and columns
- Multiple types of observational units are stored in the same table
- A single observational unit is stored in multiple tables
- The dataset is either too long or too wide


### Pivoting

Most data encountered in practice will be untidy. This is because most
people are not familiar with the principles of tidy data, and data is
often organised to facilitate uses other than analysis, such as making
entry easier.


Two common problems to look for are:

- One variable might be spread across multiple columns
- One observation might be scattered across multiple rows

Usually, a dataset will only suffer from one of these problems.

To resolve them, the `tidyr` package provides two key functions:

- `pivot_longer()`
- `pivot_wider()`

These functions are illustrated with example datasets included in the
`tidyr` package. The tables (`table2`, `table4a`) contain data on the
number of tuberculosis (TB) cases recorded in different countries for
the years 1999 and 2000. The variable `cases` represents the number of
TB cases reported for a given country, year, and type of measure.

#### Pivot longer

A common problem is a dataset where some of the column names are not
variable names, but values of a variable:

```{r}
table4a
```

To tidy a dataset like this, pivot the offending columns into a new pair
of variables.

Steps:

- Select the columns whose names are values, not variables. In this
  example, those are 1999 and 2000.
- Choose the variable to move the column names to (here, `year`).
- Choose the variable to move the column values to (here, `cases`).

```{r}
table4a |> 
  pivot_longer(c(`1999`, `2000`), names_to = "year", values_to = "cases")
```


In the final result, the pivoted columns are dropped, and new year and
cases columns are created. Other variables, such as country, are
preserved. The `cases` column now explicitly records the number of TB
cases for each year and country.


### Pivot wider

Function `pivot_wider()` is the opposite of `pivot_longer()`. You use
it when an observation is scattered across multiple rows.

```{r}
table2
```

To tidy this up, analyze the representation in a similar way to
`pivot_longer()`:

- The column to take variable names from (here, type).
- The column to take values from (here, count).

```{r}
table2 |>
  pivot_wider(names_from = type, values_from = count)
```


In this result, values of `type` (cases and population) become separate
columns, and their associated numbers from `count` fill in the values.
This produces a clearer dataset where each row corresponds to a country
and year with distinct variables for cases and population.


### Separating


The `separate()` function is used to pull apart one column into multiple
columns by splitting wherever a separator character appears. This is
useful when a single column actually contains more than one variable.

Consider the dataset `table3` included in the `tidyr` package:

```{r}
table3
```

Notice the `rate` column. It contains two variables combined into a
single column: the number of cases and the population size, separated by
a forward slash. To make the dataset tidy, these should be split into
separate variables.



The `separate()` function takes the name of the column to split and the
names of the new columns to create:

```{r}
table3 %>%
separate(rate, into = c("cases", "population"))
```

This produces two new columns, `cases` and `population`, replacing the
original `rate` column. The new columns now contain integer values for
the reported tuberculosis cases and the population in each country and
year.


By default, `separate()` splits values wherever it sees a non-alphanumeric
character, meaning any character that is not a number or letter. In the
example above, it automatically detected and split at the forward slash.


If you want to be explicit, you can specify the character to split on
with the `sep` argument:

```{r}
table3 %>%
separate(rate, into = c("cases", "population"), sep = "/")
```

This ensures that the column is split exactly where expected, giving you
clearer control over the separation process.


### Unite

The `unite()` function is the inverse of `separate()`. It combines
multiple columns into a single column. This can be useful when two or
more variables are stored in separate columns but logically belong
together.

Consider the dataset `table5`:

```{r}
table5
```

In this table, the year of observation is split into two columns,
`century` and `year`. To make the dataset easier to work with, we can
combine these into a single column.

```{r}
table5 %>% 
  unite(new, century, year)
```

By default, `unite()` places an underscore (`_`) between the values from
different columns. In this case, that would produce values like
`19_99`.

If we want the numbers to run together without any separator, we can
control this with the `sep` argument:

```{r}
table5 %>% 
  unite(new, century, year, sep = "")
```

This produces a single column `new` with values such as `1999` and
`2000`, giving a cleaner representation of the year variable.
