---
title: "{{< var assignment_title >}}"
author:
  - name: "{{< var student_name >}}"
    affiliation: "{{< var course_code >}} — {{< var course_title >}}"
date: "{{< var submission_date >}}"
format:
  html: default
  pdf: default

# Student-editable variables (fill these in before rendering)
# Tip: keep the curly braces intact; edit only the text on the right-hand side.
# Example: student_name: "Avery Lee"
student_name: "Your Name"
student_id: "Your Student ID"
assignment_title: "HW 01: Getting Started"
submission_date: "2025-09-01"
collaborators: "None"
hours_spent: "2"
# Optional assignment-specific variables
due_date: "2025-09-03"
honor_pledge_signed: "Yes"

# Per-document options
editor: visual
toc: true
number-sections: true
---

::: {.callout-note}
**How to use this template**  
1. Fill in the variables in the YAML header above (student name, ID, etc.).  
2. Knit/render with **Quarto**: `quarto render` (HTML default; PDF if LaTeX is installed).  
3. Submit the **rendered output** (`homework.html` or `homework.pdf`) and this source file `homework.qmd` unless told otherwise.
:::

## 1. Honor & Collaboration

I affirm that this submission is my own work. I have listed any collaborators and external resources used.  
**Honor pledge signed:** {{< var honor_pledge_signed >}}

- **Collaborators:** {{< var collaborators >}}
- **Hours spent:** {{< var hours_spent >}}

## 2. Short Answers

### 2.1
Type your answer here. Use complete sentences.

### 2.2
You can typeset math like $\hat\beta=(X^\top X)^{-1}X^\top y$ using standard LaTeX.

### 2.3 Citations
Cite readings, websites, or datasets, e.g., @wickham2016r. Your references will appear at the end.

## 3. R Code

Below is a minimal R example. Run it, show outputs, and explain what you see.

```{r}
#| label: setup
#| message: false
#| warning: false
# This chunk can be used for package setup.
# Example:
# install.packages("tidyverse")  # (uncomment if needed)
library(datasets)

summary(iris)
```

Describe your results in words.

### 3.1 A Plot

```{r}
#| fig-cap: "Sepal length vs sepal width"
plot(iris$Sepal.Length, iris$Sepal.Width)
```

Briefly interpret the figure.

## 4. (Optional) Python Code

If you have Python set up in Quarto, you can also include Python chunks.

```{python}
sum([1, 2, 3, 4, 5])
```

## 5. Checklist (delete after use)

- [ ] All YAML variables filled (name, ID, title, dates).
- [ ] Every question answered in complete sentences.
- [ ] Code runs top-to-bottom without errors.
- [ ] Explanations accompany tables/figures.
- [ ] Final render produced (HTML/PDF as requested).
- [ ] File(s) named correctly per submission instructions.

## 6. Appendix

Add any additional notes, derivations, or extra credit here.

## References